A new method to estimate dark matter halo concentrations

### 2015-10-31 paper/paper.tex
comparison plots inside manuscript
version con comentarios de Juan Carlos

We present a new method to estimate the concentration of dark matter
halos in N-body simulations.
Our method finds the concentration value to match integrated mass profile as a function of
halo radius.
The main advantage of this method is that it uses the full particle informatio
without any binning.
We test our method both on mock and N-body halos to compare it against
two popular methods to find concentrations: maximum radial velocity
measurements and radial particle binning to estimate the density.
Tests on the mock halos show that the accuracy of our method to
recover input concentrations varies with the number of particles in
the halo. 
For halos sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy, while for the maximumradial
velocity and density methods the accuracy is on the order of $20\%$
and $100\%$, respectively.  
For halos with $2\times 10^4$ particles our method achieves an accuracy of
$0.01\%$ while the velocity and  methods achieve $0.1\%$ and $1\%$,
respectively.  
We also measure the mass-concentration relationship on N-body data
taking care of using halos sampled with at least $200$ particles..
We find that at low masses $10^{12}\hMsun < M<10^{13}\hMsun$ our
method yields median concentration values lower by a $15\%-20\%$
compared to the velocity and density methods. 
At higher masses $3\times 10^{13}\hMsun < M < 2\times 10^{14}\hMsun$ the
three methods give similar results.
\end{abstract}

@@

On galactic scales the most striking result is that dark matter
overdensities closely follow a universal density profile. 
In a first approximation this profile is spherically symmetric and its
density only dependens on the radial coordinate.
The universality of this profiles seems to be independent of the
cosmological parameters and is self-similar for different spatial
scales after an adequate re-scaling is applied. 
\citep{NFW,Taylor2001}

One the most popular parameterization for a dark matter halo radial
density distribution is the Navarro-Frenk-White (NFW) profile
\citep{NFW}.
This profile is a double power law in radius, where the transition break
happens at the so-called scale radius, $r_s$.
The ratio between the scale radius and the virial radius $R_v$, which
defines a natural scale for the halo, is known as the concentration
$c=R_v/r_s$. 
Simulations also show that the concentration is a strong function of
halo mass and redshift.

However, high resolution simulations of Milky Way sized dark matter halos
\citep{Navarro2010} show that the universality property is not
perfect and that a better fitting parameterization to the radial
density profile is provided by the Einasto profile
\citep{Einasto1965}.
Nevertheless, the NFW density profile and its concentration have
become a standard metric to describe the structure of dark matter
halos.

@@

The second method solves the two problems mentioned above.
It works with low particles numbers and does not involve data
binning.
However, it effectively takes into account only a single data point
and discards the behaviour of the ratio $V_{\rm circ}/V_{\rm vir}$ below and
above its maxima.
Additionally, small fluctuations on the value of this maximum can
yield large perturbations on the estimated concentration parameter.

In this paper we propose a new method to estimate the dark matter halo
concentration in halos from N-body simulations.
Our method builds the cumulative mass profile from the particle
to find the best possible concentration value fitting this
distribution. 
This proposal has two advantages with respect to the two methods mentioned
above. 
It does not involve any data binning and does not throw away data points.

@@

Let us review first the basic properties of the NFW density
profile. 
This help us to define our notation and review the important
assumptions of the two different methods commonly used to measure the
concentration to finally present the new method.

@@

\begin{equation}
\rho(r) = \frac{\rho_c\delta_c}{r/r_s(1+r/r_s)^2},
\label{eq:definition}
\end{equation}
%
where $\rho_c\equiv 3H^2/8\pi G$ is the Universe critical density
constructed from the Hubble constant $H$ and the universal
gravitational constant $G$,
$\delta_c$ is the halo dimensionless characteristic density and $r_s$
is the scale radius. This radius marks the transition
between the power law scaling $\rho\propto r^{-1}$ for
$r<r_s$ and $\rho\propto r^{-3}$ for  $r>r_s$.

We define the virial radius of a halo, $r_v$, as the boundary of the
spherical volume that encloses a density of $\Delta_h$ times
the average density of the Universe. 
The corresponding mass $M_{v}$, the virial mass, can be written as
$M_{v} = \frac{4\pi}{3}\bar{\rho}\Delta_h r_v^3$.
From these virial quantities we define new dimensionless variables for
the radius and mass $x\equiv r/r_v$ and $m\equiv
M(<r)/M_v$.

In this paper we use $\Delta_h=740$ roughly corresponding to 200 times
the critical density.

\subsection{Integrated mass profile}
From these definitions we can compute the total mass enclosed inside a
radius $r$:
\begin{equation}
M(<r) = 4\pi\rho_c\delta_c  r_s^3\left[\ln \left
  (\frac{r_s+r}{r_s}\right) - \frac{r}{r_s+r}\right],
\end{equation}
%
or in terms of the dimensionless mass and radius variables

@@

%In Figure \ref{fig:profiles} we show $m(<x)$ as
%a function of $x$ for different values of the concentration in the range
%$1\leq c \leq 20$.

@@

\subsection{Estimates from the density and velocity profiles}
There are two standard methods to estimate concentrations in dark matter
halos extracted from N-body simulations.

The first method tries to directly estimate the density profile.
It takes all the particles in the halo and bins them in the logarithm
of the radial coordinate from the halo center.
Then, it estimates the density in each logarithmic bin counting the
particles and dividing by the corresponding shell volume.
At this point is is possible to make a direct fit to the density as a
function of the radial coordinate.
This method has been most recently used by \cite{Ludlow2014} to study
the mass-concentration-redshift relation of dark matter halos using
the Millennium Simulation Series.

@@

We divide the enclosed mass mass $M_i$ and the radii $r_i$ by their
virial values to obtain the dimensionless variables $m_i$ and $x_i$.
Once the mass profile is expressed in dimensionless variables the
concentration is the only free parameter. We use an Affine Invariant
Markov chain Monte Carlo implemented in the python module emcee
\citep{emcee} to sample the likelihood function distribution defined
by ${\cal L}(c)\propto \exp(-\chi^2(c)/2)$ where the $\chi^2(c)$ is written as

@@

We generate in total $400$ mock halos split into four different
groups of $100$ halos each.
The four groups differ in the total number of particles for their halos:
$20$, $200$, $2000$ and $20000$.
Inside each group the halos have random concentration values in
the range $1<c<20$ with a uniform distribution.
For all these halos ww the concentration values using the
density, velocity and mass methods described in the previous
section. 
We quantify the difference between the expected $c_{in}$
and obtained $c_{out}$ values by
%
\begin{equation}
D=(c_{in}-c_{out})/c_{in},
\label{eq:D}
\end{equation}
%
and
%
\begin{equation}
\avg{|D|}=\frac{1}{\left|{\cal{H}}\right|}\sum_{{\cal{H}}} |D|,
\end{equation}
%
where ${\cal{H}}$ corresponds to a set of mock halos, 
and $\left|{\cal{H}}\right|$ is the number of haloes in
${\cal{H}}$.

\begin{figure*}
\begin{center}
  \includegraphics[width=0.70\textwidth]{error.pdf}
\end{center}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. Our method provide the most accurate estimate at fixed
  particle number $N$.
    \label{fig:error}}
\end{figure*}

@@

The method based on the maximum of the circular velocity shows a similar
behaviour $\avg{|D|}\propto N^{-1/2}$. 
Its accuracy is $2-5$ times less
than in our method, on the order of $20\%$ for $20$ particle halos and
$0.5\%$ for $20000$ particle halos. 
The method based on the direct density
fit shows the lowest accuracy for a low particle number and an intermediate
accuracy between the other two methods for a high particle number.

@@

\label{sec:data}
\begin{figure*}
  \begin{center}
    \includegraphics[width=0.33\textwidth]{mass-velocity.pdf}
    \includegraphics[width=0.33\textwidth]{mass-density.pdf}
    \includegraphics[width=0.33\textwidth]{density-velocity.pdf}
  \end{center}
  \caption{Comparison between the concentrations measured by our
    method and the maximum velocity (left) and density (middle)
    methods. The line indicates the equal value between the two
    techniques. The right panel compares the results of the maximum
    velocity and density methods.
  \label{fig:mdv}}
\end{figure*}

@@

We build a sample of all halos located in a cubic sub-volume of $100$
\hMpc on a side centered on the most massive halo in the simulation at
$z=0$ which corresponds to the \texttt{miniMDR1} tables in the
database.
From this sample we select all the halos at $z=0$ detected with a
Friends-of-Friends (FoF) algorithm with masses in the interval
$10^{12}\leq M_{\rm FoF}/\hMsun \leq 10^{15}$.
The FoF algorithm used with a linking length of $0.17$ times the average
inter-particle distance. This choice translates into an overdensity
$\Delta_h\sim 400-700$ dependent on the halo concentration
\citep{More2011}.
Finally, we select from the database all the particles that belong to
each FoF group.

@@

>>>>

We present a new method to estimate the concentration of dark matter
halos in N-body simulations.  Our method finds the concentration value
to match {\bf the} integrated mass profile as a function of halo
radius.  The main advantage of this method is that it uses the full
particle information without any binning, {\bf which makes it very
  switable in cases when resolution becomes important}. We test our
method both on mock and N-body halos to compare it against two popular
methods to find concentrations: maximum radial velocity measurements
and radial particle binning to {\bf fit the density profile}. Tests on
the mock halos show that the accuracy of our method to recover input
concentrations varies with the halo {\bf resolution}. For halos
sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy. {\bf For the maximum radial
  velocity method it is on the order of $20\%$ while density profile
  methods perform worst at the low resolution limit since they require
  at least 200 particles inside the virial radius to make a reliable
  estimate of the density profile}. For halos with $2\times 10^4$
particles our method achieves an accuracy of $0.01\%$ while the
velocity and density profile methods achieve $0.1\%$ and $1\%$,
respectively. We measure the mass-concentration relationship on N-body
data taking care of using halos sampled with at least $200$
particles. We find that at low masses $10^{12}\hMsun <
M<10^{13}\hMsun$ our method yields median concentration values lower
by a $15\%-20\%$ compared to the velocity and density methods ({\bf
  why?}).  At higher masses $3\times 10^{13}\hMsun < M < 2\times
10^{14}\hMsun$ the three methods give similar results.
\end{abstract}

@@

{\bf Navarro et.al. 2010 have used high resolution simulations to show
  that the full halo density profile is not universal and follows an
  Einasto density distribution. However, if one is not interested in
  the very central region of the dark matter halo (where galaxy
  formation takes place, and where the effects of baryon physics on
  dark matter distribution is still an uncknown), one can assume an
  approximation where dark matter overdensities closely follow an
  universal density profile. In a first approximation this profile is
  spherically symmetric and its density only dependens on the radial
  coordinate. Although it is well known that the characteristics of
  this densty profile depends on redshift (Maccio et.al. 2008), its
  universal shape seems to be independent of the cosmological
  parameters and is self-similar for different spatial scales after an
  adequate re-scaling is applied. \citep{NFW,Taylor2001}
}

{\bf In such circumstances}, one the most popular parameterizations
for dark matter halo radial density distribution is the
Navarro-Frenk-White (NFW) profile \citep{NFW}.  This profile is a
double power law in radius, where the transition break happens at the
so-called scale radius, $r_s$.  The ratio between the scale radius and
the virial radius $R_v$ is known as the concentration $c=R_v/r_s$.
Simulations show that the concentration is a strong function of halo
mass and redshift and provides information about the mass assembly of
the dark matter halo (Citar paper que reproduce mis graficas y conecta
C con m(z)).

@@

The second method solves the two problems mentioned above.  It works
with low particle numbers and does not involve data binning.  However,
it effectively takes into account only a single data point and
discards the behaviour of the ratio $V_{\rm circ}/V_{\rm vir}$ below
and above its maxima.  Additionally, small fluctuations on the value
of this maximum can yield large perturbations on the estimated
concentration parameter.

In this paper we propose a new method to estimate the dark matter halo
concentration in halos from N-body simulations.  Our method builds the
cumulative mass profile from the particle to find the best possible
concentration value fitting this distribution.  This proposal has two
advantages with respect to the two methods mentioned above.  It does
not involve any data binning and does not throw away data points {\bf
  resulting in a robust estimate even at low resolution/particle
  numbers. In the same sence, since the method does not require any
  binning, there is no need to tune numerical parameters, which
  provide definitive/independent estimates for the concentration
  parameter.}.

{\it mi opinión es que el método es muy bueno, y sobresalta sobre los
  otros dos en sus ventaja: robusto, no requiere bineo, aplica bien en
  resoluciones pobres. Como herramienta (que es lo que se está
  vendiendo en el trabajo, un método, eso es lo que hay que recalcar y
  resaltar)}

@@

Let us review first the basic properties of the NFW density profile.
This shall help us to define our notation and review the important
assumptions of the two different methods commonly used to measure the
concentration to finally present the new method.

@@

\begin{equation}
\rho(r) = \frac{\rho_c\delta_c}{r/r_s(1+r/r_s)^2},
\label{eq:definition}
\end{equation}
%
where $\rho_c\equiv 3H^2/8\pi G$ is the Universe critical density, $H$
is the Hubble constant, $G$ and the universal gravitational constant,
$\delta_c$ is the halo dimensionless characteristic density and $r_s$
is the scale radius. This radius marks {\bf the point where the
  logarithmic slope of the density profile is equal to -2} marking the
transition between the power law scaling $\rho\propto r^{-1}$ for
$r<r_s$ and $\rho\propto r^{-3}$ for $r>r_s$.

We define the virial radius of a halo, $r_v$, as the boundary of the
spherical volume that encloses a density of $\Delta_h$ times the mean
density of the Universe.  The corresponding mass $M_{v}$, the virial
mass, can be written as $M_{v} = \frac{4\pi}{3}\bar{\rho}\Delta_h
r_v^3$.  From these virial quantities we define new dimensionless
variables for the radius and mass $x\equiv r/r_v$ and $m\equiv
M(<r)/M_v$.

In this paper we use $\Delta_h=740$ roughly corresponding to 200 times
the critical density {\bf at redshift z=0}.

\subsection{Integrated mass profile}

From these definitions we can compute the total mass enclosed inside a
radius $r$:
\begin{equation}
M(<r) = 4\pi\rho_c\delta_c  r_s^3\left[\ln \left
  (\frac{r_s+r}{r_s}\right) - \frac{r}{r_s+r}\right],
\end{equation}
%
or in terms of the dimensionless mass and radius variables

@@

@@

\subsection{Estimates from the density and velocity profiles}
{\bf To date,} there are two standard methods to estimate
concentrations in dark matter halos extracted from N-body simulations.

The first method tries to directly estimate the density profile.  It
takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  Then, it estimates the
density in each logarithmic bin counting the particles and dividing by
the corresponding shell volume.  At this point is is possible to make
a direct fit to the density as a function of the radial coordinate.
This method has been {\bf used broadly for more than two decades} to
study the mass-concentration-redshift relation of dark matter halos
using the Millennium Simulation Series.

{\it no veo la necesidad de citar el ``trabajo mas'' reciente que usa
  el metodo, cuando ha sido usado durante mas de dos decadas!}

@@

We divide the enclosed mass mass $M_i$ and the radii $r_i$ by their
virial values to obtain the dimensionless variables $m_i$ and $x_i$.
Once the mass profile is expressed in dimensionless variables the
concentration is the only free parameter. We use an Affine Invariant
Markov chain Monte Carlo implemented in the python module {\em emcee}
\citep{emcee} to sample the likelihood function distribution defined
by ${\cal L}(c)\propto \exp(-\chi^2(c)/2)$ where the $\chi^2(c)$ is
written as

@@

We generate in total $400$ mock halos split into four different groups
of $100$ halos each.  The four groups differ in the total number of
particles for their halos: $20$, $200$, $2000$ and $20000$.  Inside
each group the halos have random concentration values in the range
$1<c<20$ with a uniform distribution.  For all these halos {\bf ww?}
the concentration values using the density, velocity and mass methods
described in the previous section.  We quantify the difference between
the expected $c_{in}$ and obtained $c_{out}$ values by
%
\begin{equation}
D=(c_{in}-c_{out})/c_{in},
\label{eq:D}
\end{equation}
%
and
%
\begin{equation}
\avg{|D|}=\frac{1}{\left|{\cal{H}}\right|}\sum_{{\cal{H}}} |D|,
\end{equation}
%
where ${\cal{H}}$ corresponds to a set of mock halos, and
$\left|{\cal{H}}\right|$ is the number of haloes in ${\cal{H}}$. {\bf
  The larger the value of $\avg{|D|}$, the larger the discrepace of
  the concentrations obtained for that method}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.70\textwidth]{error.pdf}
\end{center}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. Our method provide the most accurate estimate at fixed
  particle number $N$.  {\it Con respecto a esta figura tengo una
    pregunta. Yo ajusté los perfiles con menos de 200 partulas? no
    recuerdo!}
    \label{fig:error}}
\end{figure*}

@@

The method based on the maximum of the circular velocity shows a
similar behaviour $\avg{|D|}\propto N^{-1/2}$.  Its accuracy is $2-5$
times less than in our method, on the order of $20\%$ for $20$
particle halos and $0.5\%$ for $20000$ particle halos. {\bf The method
  based on the direct density fit shows the lowest accuracy for a low
  particle number. As it is discussed in \citep{Munoz2011}, density
  binning for halos with particle numbers below $\sim200$ leads to a
  biased estimation of the mass density profile, and therefore to a
  biased estimation of the concentration parameter. This behaviour is
  evident in the large values of $\avg{|D|}$ for halos with number of
  particles below ~200 and an intermediate accuracy between the other
  two methods for a high particle number.}

@@

\label{sec:data}
\begin{figure*}
  \begin{center}
    \includegraphics[width=0.33\textwidth]{conc_mass_vel.pdf}
    \includegraphics[width=0.33\textwidth]{conc_mass_dens.pdf}
    \includegraphics[width=0.33\textwidth]{conc_dens_vel.pdf}
  \end{center}
  \caption{Comparison between the concentrations measured by our
    method and the maximum velocity (left) and density (middle)
    methods. The line indicates the equal value between the two
    techniques. The right panel compares the results of the maximum
    velocity and density methods.
  \label{fig:mdv}}
\end{figure*}

{\bf The results presented in the previous section show that in an
  idealized setup (perfect spherical halos, pure NFW desity profiles,
  total isolation) our method of the mass profile performs better than
  the other two methods commonly used to measure concentrations in
  simulated dark matter haos. However, actual simulated dark matter
  halos are much more complex. They are not isolated, many of them
  experience mergers that take them out of the equilibrium. Halos may
  have also plenty of substructure, etc. All these effects disturb the
  mass density profile, taking it away from the ideal NFW profile we
  assume in this wok. In order to see the real performance of our new
  method use it to estimate the concentration parameter in dark matter
  halos obtained from a cosmological Nbody simulation and compare with
  the other two calsical methods.}

@@

We build a sample of all halos located in a cubic sub-volume of $100$
\hMpc on a side centered on the most massive halo in the simulation at
$z=0$ which corresponds to the \texttt{miniMDR1} tables in the
database.  From this sample we select all the halos at $z=0$ detected
with a Friends-of-Friends (FoF) algorithm with masses in the interval
$10^{12}\leq M_{\rm FoF}/\hMsun \leq 10^{15}$.  The FoF algorithm used
with a linking length of $0.17$ times the {\bf mean} inter-particle
distance. This choice translates into an overdensity $\Delta_h\sim
400-700$ dependent on the halo concentration \citep{More2011}.
Finally, we select from the database all the particles that belong to
each FoF group.

@@

{\em Aqui hay algo importante que hay que entender antes de discutir
  esta gráfica , y es la razón por la cual el metodo del perfil de
  masa da valores de concentración tan bajos a baja masa y comparables
  a los de los otros dos métodos a alta masa. Es importante entender
  la razón de esto!.  Se me ocurren unas preguntas para tratar de
  entender. \\ 1) La grafica de c-M \ref{fig:concentration} usa todos
  los halos o solo los halos relajados? La inclusión de halos no
  relajados introduce in bias en la mediana en Cvir, así que vale la
  pena inclusive agregar una linea mas a la gráfica con los halos
  relajados y sin ellos. Para los otros dos métodos ya se sabe que
  hace la población de halos no relajados en la medida así que no creo
  necesario agregarlas para esos métodos. \\ Hay que explicar un poco
  mas la gráfica en el sentido de que se grafica y como se
  calculó. mean concentration?, Median concentrations? o mean in
  logscale?  Mas bins de cuanto? Mínimo cuantos halos por mass bin?,
  Halos relajados en todos los casos, o indistintos?}
